Generating Images From Captions\\ With Attention

### 2015-10-30 iclr-paper-new.tex
update

\cite{kingma_vae} have introduced the Variational Auto-Encoder (VAE) which can be seen as a neural network with continous latent variables. The encoder is used to approximate a posterior distribution and the decoder is used to stochastically reconstruct the data from latent variables. The model's efficient inference and learning procedures allow it to scale to large datasets. 
\cite{gregor_draw} introduced the Deep Recurrent Attention Writer (DRAW), extending the VAE approach by incorporating a novel differentiable attention mechanism which significantly improved performance.% as well as quality of generated samples.
%While most of the samples from VAE and DRAW models resemble a clear structure of objects, the generated images are blurry most of the time.

@@

Unlike the original DRAW network where latent variables are independent unit Gaussians $\mathcal{N}(0, I)$, the latent variables in the proposed alignDRAW model have their mean and variance depend on the previous recurrent hidden states $\hdec_{t-1}$ as in \citep{bachman_sdm}, except for $\lat_1 \sim \prior(\Lat_1) = \mathcal{N}(0, I)$. It has been shown in \citep{bachman_sdm} that model performance is improved by including dependencies between latent variables. Namely, the mean and variance of $\prior(\Lat_t)$ is: 
\begin{align}
\mu_{t}(\hdec_{t-1}) &= tanh(W_{\mu}\hdec_{t-1})\\
\sigma_{t}(\hdec_{t-1}) &= exp(tanh(W_{\sigma}\hdec_{t-1})) 
\end{align}
Formally, the image is generated by iteratively computing the following equations for $t=1,...,T$ (see Figure~\ref{fig:figmodel}):
\begin{align}
\label{eq:x_hat}
\lat_t &\sim \prior(\Lat_t|\Lat_{1:t-1}) = \mathcal{N}(\mu_{t}(\hdec_{t-1}), \sigma_{t}(\hdec_{t-1})),\\
s_{t} &= align(\hdec_{t-1}, \hlangall), \\
\hdec_t &= \decoder(\hdec_{t-1}, [z_t, s_{t}]), \\
\canv_t &= \canv_{t-1} + \writeop(\hdec_t), 
\label{eq:write}
%\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}
where $\mathit{LSTM}$ is a standard \textit{Long Short-Term Memory} architecture \citep{hochreiter_lstm} with \textit{forget} gates \citep{gers_forget}, $[v_1, v_2]$ is the concatenation of the vectors $v_1$ and $v_2$ into the single vector, and $\canv_0$, $\hdec_0$ are initialized to the the learned biases. The $\writeop$ operator applies an array of Gaussian filters on the patch that places it on the canvas with smoothly varying location and zoom (same as in \citep{gregor_draw}). The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps \citep{bahdanau_mt}. Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $align$ operator outputs a dynamic sentence representation $s_t$ at each step through a weighted sum using alignment probabilities $\alpha_{1...N}^{t}$:
%\begin{align}
%\vspace{-0.1in}
$s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}^{t}\hlang_{1} + \alpha_{2}^{t}\hlang_{2} + ... + \alpha_{N}^{t}\hlang_{N}.$
%\end{align}
The corresponding alignment probabilities $\alpha_{1...N}^{t}$ at each step are obtained using the caption representation $\hlangall$ and the hidden state of the generative model $\hdec_{t}$, by first calculating energies using a $tanh$ perceptron scaled by vector $v$ and then normalizing them, as in \citep{bahdanau_mt}:
%\begin{align}
%\vspace{-0.1in}
$\alpha_{k}^{t}=\frac{exp(v^{T}tanh(U\hlang_{k} + W\hdec_{t} + b)}{\sum_{j=1}^{N}exp(v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)}.$
%\end{align}
The final canvas $\canv_T$ is transformed using a sigmoid function to produce a conditional Bernoulli data likelihood over pixels given the latent variables and the input caption.

\comm{
We extended the architecture of the DRAW network to include an additional input caption from the language model described in Sec. (\ref{sec:lang}). Similarly to the original DRAW network, the conditional DRAW network is a stochastic recurrent neural network that consists of an Inference LSTM that infers the distribution of latent variables of image $x$ given caption $y$ and then a Generative LSTM that uses the inferred latent variables in order to reconstruct the image $x$ given caption $y$, where $\icaption$ is the input caption consisting of $N$ words $y_{1}, y_{2}, ..., y_{n}$. The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps as in \cite{bahdanau_mt}:

@@

%The model is learned by the modified version of Stochastic Gradient Variation Bayes (SGVB) algorithm introduced by \cite{kingma_vae}. 
The model is trained to maximize a variational lower bound $\loss$ on the marginal likelihood of the correct image $\oimage$ given the input caption $\icaption$, where $\loss = \sum_{\Lat}Q(\Lat\given\oimage,\icaption)\left(\log P(\oimage\given\icaption, \Lat_{1:T}) + \kldiv\left(Q(\Lat\given\oimage,\icaption)\klBars P(\Lat)\right)\right) \le \log P(\oimage\given\icaption)$. %The posterior inference over the latent variables $\Lat_{1:T}$ is approximated using the inference RNN. 
Similarly to the DRAW network, the inference RNN produces an approximate posterior $Q(\Lat_{1:T}\given\oimage,\icaption)$ via a $\readop$ operator, which reads a patch from image using an array of Gaussian filters (inverse of $\writeop$), over the training image at each step. 
\begin{align}
\label{eq:x_hat}
\hat{\oimage}_t &= \oimage-\sigmoid(\canv_{t-1})\\
\label{eq:read}
r_t &= \readop(\oimage_t, \hat{\oimage}_t, \hdec_{t-1})\\
\henc_t &= \encoder(\henc_{t-1}, [r_t, \hdec_{t-1}])\\
\post(\Lat_t|\oimage,\icaption,\Lat_{1:t-1}) &= \mathcal{N}\left(\mu_t(\henc_t), \sigma_t(\henc_t)\right)
\end{align}

@@

where $\hat{\oimage}$ is the \textit{error} image, $\sigmoid$ operator is a logistic sigmoid function $\sigmoid=\frac{1}{1+exp(-x)}$ and $\henc_0$ is initialized to the learned bias.

The terms in the variational lower bound are rearranged using the law of total expectation. Therefore, the loss function $\loss$ is calculated as follows:
\begin{align}
\loss =  &\expectation_{Q(\Lat_{1:T}\given\icaption,\oimage)}\left[ - \log p(\oimage\given\icaption,\Lat_{1:T}) + \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\Lat_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\Lat_{t-1},\icaption)\right) \right] \nonumber \\& + \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
The expectation can be approximated by $\numSamples$ Monte Carlo samples $\LatSample_{1:T}$ from $\post(\Lat_{1:T}\given\icaption, \oimage)$:
\begin{align}
\loss \approx  &\frac{1}{\numSamples}\sum_{\sampleIdx=1}^\numSamples\left[ - \log p(\oimage\given\icaption,\LatSample^\sampleIdx_{1:T}) + \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption)\right) \right] \nonumber \\& + \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
We used only one sample during training for all our experiments.
\comm{
\begin{align}
\loss &= -\sum_{t=1}^{T}D_{KL}(\post(\Lat_t|\henc_t,s_{t-1})\,||\,\prior(\Lat_t)) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)\\
&=
\frac{1}{2}\sum_{t=1}^{T}(1 - 2\,log\,\sigma_{t}^{prior} + 2\,log\,\sigma_{t} - \frac{exp(2\,log\,\sigma_{t}) + (\mu_{t} - \mu_{t}^{prior})^{2}}{exp(2\,log\,\sigma_{t}^{prior})}) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)
\end{align}
}

>>>>

\cite{kingma_vae} have introduced the Variational Auto-Encoder (VAE) which can be seen as a neural network with continuous latent variables. The encoder is used to approximate a posterior distribution and the decoder is used to stochastically reconstruct the data from latent variables. The model's efficient inference and learning procedures allow it to scale to large datasets. 
\cite{gregor_draw} introduced the Deep Recurrent Attention Writer (DRAW), extending the VAE approach by incorporating a novel differentiable attention mechanism which significantly improved performance.% as well as quality of generated samples.
%While most of the samples from VAE and DRAW models resemble a clear structure of objects, the generated images are blurry most of the time.

@@

Unlike the original DRAW network where latent variables are independent unit Gaussians $\mathcal{N}(0, I)$, the latent variables in the proposed alignDRAW model have their mean and variance depend on the previous recurrent hidden states $\hdec_{t-1}$, except for $\lat_1 \sim \prior(\Lat_1) = \mathcal{N}(0, I)$.  Namely, the mean and variance of prior distribution $\prior(\Lat_t)$ is parameterized by: 
\begin{align}
\mu_{t}(\hdec_{t-1}) &= tanh(W_{\mu}\hdec_{t-1}),\\
\sigma_{t}(\hdec_{t-1}) &= {exp}\big(tanh(W_{\sigma}\hdec_{t-1})\big),
\end{align}
where $W_{\mu}, W_{\sigma}$ are learnt model parameters. It has been shown in \citep{bachman_sdm} that model performance is improved by including dependencies between latent variables. Formally, the image is generated by iteratively computing the following equations for $t=1,...,T$ (see Figure~\ref{fig:figmodel}):
\begin{align}
\label{eq:x_hat}
\lat_t &\sim \prior(\Lat_t|\Lat_{1:t-1}) = \mathcal{N}\big(\mu_{t}(\hdec_{t-1}), \sigma_{t}(\hdec_{t-1})\big),\\
s_{t} &= align(\hdec_{t-1}, \hlangall), \\
\hdec_t &= \decoder(\hdec_{t-1}, [z_t, s_{t}]), \\
\canv_t &= \canv_{t-1} + \writeop(\hdec_t), 
\label{eq:write}
%\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}
where $\mathit{LSTM}$ is a standard \textit{Long Short-Term Memory} architecture \citep{hochreiter_lstm} with \textit{forget} gates \citep{gers_forget}, $[v_1, v_2]$ is the concatenation of the vectors $v_1$ and $v_2$ into the single vector, and $\canv_0$, $\hdec_0$ are initialized to the the learned biases. The $\writeop$ operator applies an array of Gaussian filters on the patch that places it on the canvas with smoothly varying location and zoom (same as in \citep{gregor_draw}). The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps \citep{bahdanau_mt}. Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $align$ operator outputs a dynamic sentence representation $s_t$ at each step through a weighted sum using alignment probabilities $\alpha_{1...N}^{t}$:
%\begin{align}
%\vspace{-0.1in}
$s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}^{t}\hlang_{1} + \alpha_{2}^{t}\hlang_{2} + ... + \alpha_{N}^{t}\hlang_{N}.$
%\end{align}
The corresponding alignment probabilities $\alpha_{1...N}^{t}$ at each step are obtained using the caption representation $\hlangall$ and the hidden state of the generative model $\hdec_{t}$, by first calculating energies using a $tanh$ perceptron scaled by vector $v$ and then normalizing them, as in \citep{bahdanau_mt}:
%\begin{align}
%\vspace{-0.1in}
$\alpha_{k}^{t}=\frac{exp(v^{T}tanh(U\hlang_{k} + W\hdec_{t} + b)}{\sum_{j=1}^{N}exp(v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)}.$
%\end{align}
The final canvas $\canv_T$ is transformed using a sigmoid function to produce a conditional Bernoulli data likelihood over the image pixels $\oimage$ given the latent variables $\Lat$ and the input caption $\icaption$,
\begin{align}
P(\oimage\given\icaption, \Lat_{1:T}) = \prod_i P(x_i\given\icaption, \Lat_{1:T}) = \prod_i \sigmoid(\canv_{T,i}),
\end{align}
where, $\canv_{T,i}$ is the $i^{th}$ pixel on the final canvas computed at step T and $\sigmoid$ operator is a logistic sigmoid function $\sigmoid(x)=\frac{1}{1+exp(-x)}$. 
\comm{
We extended the architecture of the DRAW network to include an additional input caption from the language model described in Sec. (\ref{sec:lang}). Similarly to the original DRAW network, the conditional DRAW network is a stochastic recurrent neural network that consists of an Inference LSTM that infers the distribution of latent variables of image $x$ given caption $y$ and then a Generative LSTM that uses the inferred latent variables in order to reconstruct the image $x$ given caption $y$, where $\icaption$ is the input caption consisting of $N$ words $y_{1}, y_{2}, ..., y_{n}$. The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps as in \cite{bahdanau_mt}:

@@

%The model is learned by the modified version of Stochastic Gradient Variation Bayes (SGVB) algorithm introduced by \cite{kingma_vae}. 
The model is trained to maximize a variational lower bound $\loss$ on the marginal likelihood of the correct image $\oimage$ given the input caption $\icaption$, where $\loss = \sum_{\Lat}Q(\Lat\given\oimage,\icaption)\left(\log P(\oimage\given\icaption, \Lat_{1:T}) - \kldiv\left(Q(\Lat\given\oimage,\icaption)\klBars P(\Lat)\right)\right) \le \log P(\oimage\given\icaption)$. %The posterior inference over the latent variables $\Lat_{1:T}$ is approximated using the inference RNN. 
Similarly to the DRAW network, the inference RNN produces an approximate posterior $Q(\Lat_{1:T}\given\oimage,\icaption)$ via a $\readop$ operator, which reads a patch from image using an array of Gaussian filters (inverse of $\writeop$), over the training image at each step. 
\begin{align}
\label{eq:x_hat}
\hat{\oimage}_t &= \oimage-\sigmoid(\canv_{t-1}),\\
\label{eq:read}
r_t &= \readop(\oimage_t, \hat{\oimage}_t, \hdec_{t-1}),\\
\henc_t &= \encoder(\henc_{t-1}, [r_t, \hdec_{t-1}]),\\
\post(\Lat_t|\oimage,\icaption,\Lat_{1:t-1}) &= \mathcal{N}\left(\mu_t(\henc_t), \sigma_t(\henc_t)\right),
\end{align}

@@

where $\hat{\oimage}$ is the \textit{error} image and $\henc_0$ is initialized to the learned bias.

The terms in the variational lower bound are rearranged using the law of total expectation. Therefore, the loss function $\loss$ is calculated as follows:
\begin{align}
\loss =  &\expectation_{Q(\Lat_{1:T}\given\icaption,\oimage)}\left[ \log p(\oimage\given\icaption,\Lat_{1:T}) - \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\Lat_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\Lat_{t-1},\icaption)\right) \right] \nonumber \\& - \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
The expectation can be approximated by $\numSamples$ Monte Carlo samples $\LatSample_{1:T}$ from $\post(\Lat_{1:T}\given\icaption, \oimage)$:
\begin{align}
\loss \approx  &\frac{1}{\numSamples}\sum_{\sampleIdx=1}^\numSamples\left[ \log p(\oimage\given\icaption,\LatSample^\sampleIdx_{1:T}) - \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption)\right) \right] \nonumber \\& - \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
We used only one sample during training for all our experiments.
\comm{
\begin{align}
\loss &= -\sum_{t=1}^{T}D_{KL}(\post(\Lat_t|\henc_t,s_{t-1})\,||\,\prior(\Lat_t)) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)\\
&=
\frac{1}{2}\sum_{t=1}^{T}(1 - 2\,log\,\sigma_{t}^{prior} + 2\,log\,\sigma_{t} - \frac{exp(2\,log\,\sigma_{t}) + (\mu_{t} - \mu_{t}^{prior})^{2}}{exp(2\,log\,\sigma_{t}^{prior})}) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)
\end{align}
}
